Symmetric Circuits for Rank Gates

### 2017-07-26 paper.tex
26/07/2017

\documentclass[12pt]{report}

% This first part of the file is called the PREAMBLE. It includes
% customizations and command definitions. The preamble is everything
% between \documentclass and \begin{document}.

\usepackage[margin=1in]{geometry} % set the margins to 1in on all sides
\usepackage{graphicx} % to include figures
\usepackage{amsmath} % great math stuff
\usepackage{amsfonts} % for blackboard bold, etc
\usepackage{amsthm} % better theorem environments

% various theorems, numbered by section

\newtheorem{thm}{Theorem} \newtheorem{claim}{Claim}
\newtheorem{remark}{Remark} \newtheorem{definition}{Definition}
\newtheorem{lem}[thm]{Lemma} \newtheorem{prop}[thm]{Proposition}
\newtheorem{cor}[thm]{Corollary} \newtheorem{conj}[thm]{Conjecture}

@@

\newcommand{\bd}[1]{\mathbf{#1}} % for bolding symbols
\newcommand{\RR}{\mathbb{R}} % for Real numbers
\newcommand{\ZZ}{\mathbb{Z}} % for Integers
\newcommand{\MB}{\mathbb{B}_{\matsym}} % for Integers
\newcommand{\SB}{\mathbb{B}_{\sym}} % for Integers
\newcommand{\col}[1]{\begin{matrix} #1 \end{matrix} \right]}
\newcommand{\comb}[2]{\binom{#1^2 + #2^2}{#1+#2}}
\newcommand{\stab}{\text{\textbf{Stab}}}
\newcommand{\setstab}{\text{\textbf{SetStab}}}
\newcommand{\matstab}{\text{\textbf{MatStab}}}
\newcommand{\inv}{\text{\textbf{inv}}}
\newcommand{\aut}{\text{\textbf{Aut}}}
\newcommand{\kernal}{\text{\textbf{ker}}}
\newcommand{\alt}{\text{\textbf{Alt}}}
\newcommand{\sym}{\text{\textbf{Sym}}}
\newcommand{\rank}{\text{\textbf{rk}}}
\newcommand{\orb}{\text{\textbf{Orb}}}
\newcommand{\SP}{\text{\textbf{SP}}}
\newcommand{\maj}{\text{\textbf{Maj}}}
\newcommand{\dom}{\text{\textbf{Dom}}}
\newcommand{\child}{\text{child}}
\newcommand{\countgate}{\text{count}}
\newcommand{\agree}{\text{AGREE}}
\newcommand{\supp}{\text{SUPP}}
\newcommand{\rankparam}{\text{PARAM}}

\newcommand{\Alpha}{A}
\newcommand{\matsym}{\text{\textbf{MatSym}}}
\begin{document}

\nocite{*}

\title{Circuits}

\author{Gregory Wilsenach \\ 
Computer Laboratory \\
University of Cambridge}

@@

\tableofcontents

\chapter{Introduction}

\chapter{Background}

\chapter{Symmetric Circuits}

\chapter{Extending Symmetric Circuits}

\chapter{Matrix-Symmetric Circuits}

The symmetric circuits of Anderson and Dawar \cite{} have been shown to express
$FPC$.

In this section we develop a notion of a matrix-symmetric circuit, an extension
of the symmetric circuit defined by Anderson and Dawar \cite{}. This is my
latest version of the support and circuit capturing result.

\section{}

\section{Circuits and Symmetry}
\begin{definition}
  Let $A$ and $B$ be sets, say that a function $f: \{ 0,1 \}^{A \times B}
  \rightarrow \{ 0,1 \}$ is matrix-symmetric if for any $\omega: A \times B
  \rightarrow \{0, 1\}$ and $(\alpha, \beta) \in \sym_A \times \sym_B$ we have
  $f((\alpha, \beta)\cdot \omega) = f (\omega)$.
  \\
  A function $f: \{0,1\}^* \rightarrow \{0,1\}$ is matrix-symmetric iff for any
  two finite sets $A$, $B$ we have that that $f|_{A \times B}$ is
  matrix-symmetric (where $f|_{A \times B}$ is the restriction of the function
  $f$ to inputs sequences of size $\vert A \times B \vert$ and where elements of
  the input sequence are labelled by elements of $A \times B$).
\end{definition}

\section{The Support Theorem}

Without a loss of generality we assume that $A$ and $B$ are always initial
segments of the natural numbers.

% Let $C$ be a circuit over some Boolean basis. For each gate $g \in C$ we
% associate a relation $\omega_g$ (or just $\omega$ when the context makes it
% clear). We say a gate is symmetric if that relation

% So far we have defined a circuit over a Boolean basis where the circuit
% elements all compute symmetric functions. The definition of an induced
% automorphism only makes. As such, we begin by augmenting the definition of a
% circuit and automorphism.

\begin{definition}[Circuits on Structures]
  For $\mathbb{B}_{\sym}$ a basis of Boolean symmetric functions,
  $\mathbb{B}_{\matsym}$ a basis of Boolean matrix-symmetric functions and
  $\tau$ a set of relation symbols, we define a $(\mathbb{B}_\sym,
  \mathbb{B}_\sym, \tau)$-circuit $C_n$ computing a $q$-ary query $Q$ is a
  structure $\langle G, W, \Omega, \Sigma, \Lambda, L\rangle$.
  \begin{itemize}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $W \subseteq G \times G$, where $W$ is called the wires of the circuit.
    $(G,W)$ must be a directed acyclic graph. For $g \in G$ we $H_g := \{ h \in
    C_n : W(h,g)\}$ be the set of children of $g$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B}_\sym \uplus
    \mathbb{B}_\matstab \uplus \tau \uplus \{0,1\} $ which maps input gates to
    $\tau \uplus \{0,1\}$ and where $\Sigma^{-1} (0) \leq 1$ $\Sigma^{-1} (1)
    \leq 1$ and the internal gates get mapped into $\mathbb{B}_\sym \uplus
    \mathbb{B}_\matstab$. Gates mapped to $\tau$ are called relational gates and
    gates mapped to 1 or 0 are called constant gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_R)_{R \in
      \tau}$ where for each $R \in \tau$, $\Lambda_R$ maps each relational gate
    $g$ with $R = \Sigma (g)$ to the tuple $\Lambda_R (g) \in [n]^r$, where $r$
    is the arity of the symbol $R$. When no ambiguity arises we write $\Lambda
    (g)$ for $\Lambda_R (g)$.
  \item $L$ is the set of labelings of $C_n$. Let $g \in G$ be an internal gate
    and $H$ be the children of $g$. Then let $S$ be a set where $\{0,1\}^S =
    \dom(\Sigma (g))$. Then $L$ is a function that assigns to $g$ an onto
    function $\omega_g:S \rightarrow H$. We call this the matrix labelling of
    $g$.
  \end{itemize}
\end{definition}

% \begin{definition}
%   let $C = \langle G, W, \Omega, \Sigma, \Lambda, L\rangle$ be a
%   $(\mathbb{B}_\sym, \mathbb{B}_\sym, \tau)$-circuit, then take $g_1,g_2 \in
%   G$ such that $\Sigma (g_1) \Sigma(g_2) \in \mathbb{B}_\matsym$.

\begin{definition}
  Let $\omega_1: A \times B \rightarrow H$ and $\omega_2: A' \times B'
  \rightarrow H'$ define the equivalence relation $\sim$ as $\omega_1 \sim
  \omega_2$ iff $\omega_1$ and $\omega_2$ have the same domain and co-domain and
  there exists $(\alpha, \beta) \in \sym_A \times \sym_B$ and for all $(a,b) \in
  A \times B$ we have $\omega_1 = \omega_2 \circ (\alpha, \beta)$.
\end{definition}

% \begin{definition}
%   Let $\omega_1: A \rightarrow H$ and $omega_2:A' \rightarrow H'$. Define the
%   equivalence relation $\sim$ as $\omega_1 \sim \omega_2$ iff $\omega_1$ and
%   $\omega_2$ have the same domain and co-domain, and
% \end{definition}

\begin{definition}[Automorphism]
  let $C = \langle G, W, \Omega, \Sigma, \Lambda, L\rangle$ be a
  $(\mathbb{B}_\sym, \mathbb{B}_\sym, \tau)$-circuit computing at $q$-ary query
  on structures of size $n$. Let $\sigma \in \sym_n$ and $\pi: G \rightarrow G$
  be a bijection such that
  \begin{itemize}
  \item for all gates $g, h \in G$, $W(g,h)$ iff $W(\pi g, \pi h)$,
  \item for all output tuples $x \in [n]^q$, $\pi \Omega (x) = \Omega (\sigma
    x)$,
  \item for all gates $g \in G$, $\Sigma (g) = \Sigma (\pi g)$, and
  \item for each relational gate $g \in G$, $\sigma \Lambda (g) = \Lambda (\pi
    g)$
  \item for each internal gate $g$ if $\Sigma (g) \in \mathbb(B)_\matsym$. Then
    we have that $L(\pi g) \sim \pi \circ L(g)$.
  \end{itemize}
\end{definition}

\begin{remark}
  I also need a new notion of robustness here in order to collapse gates that
  have the same children and equivalent labelings are properly regarded as being
  the same gate.
\end{remark}
% \begin{definition}
%   Let $C$ be a circuit, $g$ be a gate in $C$, $H$ be $g$'s set of children and
%   let $\omega: A \times B \rightarrow H$ be a surjection, where $A$ and $B$
%   are sets and $g$ takes $\vert A \vert \cdot \vert B \vert$ inputs. Call such
%   an $\omega$ a matrix labeling for $g$.
% \end{definition}

% \begin{definition}
%   Let $C$ be a circuit and $g_1$ and $g_2$ be gates in $C$. Let $\sigma \in
%   \sym_n$ be a permutation such
% \end{definition}

\begin{definition}
  Let $\omega$ be a matrix labeling for $g$. Define the matrix stabilizer for
  $\omega$, denoted by $\matstab(\omega)$, to be the set of all $\sigma \in
  \sym_n$ such that $\sigma H = H$ and there exists $(\alpha, \beta) \in \sym_A
  \times \sym_B$ such that for all $(i,j) \in A \times B$ we have that $\omega
  (\alpha i, \beta j) = \sigma \omega (i,j)$.
\end{definition}

\begin{definition}
  Let $g$ be a gate with matrix labeling $\omega$, $h,h' \in H$ and $\sigma \in
  \sym_n$. We say that a pair $(h, h')$ is compatible with $(\sigma, \omega)$ if
  $\sigma h, \sigma h' \in H$ and there exists $(\gamma_1, \gamma_2),
  (\gamma_1', \gamma_2') \in \sym_A \times \sym_B$ s.t.
  \begin{align*}
    &(\gamma_1, \gamma_2) \cdot \omega^{-1} (h) = \omega^{-1}(\sigma h), \text{ and} \\ 
    &(\gamma_1', \gamma_2') \cdot \omega^{-1} (h') = \omega^{-1}(\sigma h'),
  \end{align*}
  and for all $(i,j) \in \omega^{-1}(h)$, $(i',j') \in \omega^{-1}(h')$ we have
  that
  \begin{align*}
    &i =i' \implies \gamma_1(i) = \gamma_1'(i'), \text{ and} \\
    &j =j' \implies \gamma_2(j) = \gamma_2'(j').
  \end{align*}

\end{definition}

\begin{lem}
  Let $g$ be a gate, $\omega$ be a matrix labeling of $g$, $\sigma \in \sym_n$.
  Then $\sigma \in \matstab(\omega)$ iff for all $h,h' \in H$ we have that
  $(h,h')$ is compatible with $(\sigma, \omega)$.
\end{lem}

\begin{proof}
  $`\Rightarrow'$: We have that $\sigma \in \matstab(\omega)$ and so there
  exists $(\alpha, \beta) \in \sym_A \times \sym_B$ such that for all $(i,j) \in
  A \times B$ we have $\sigma \omega (i,j) = \omega (\alpha i, \beta j)$. From
  compatibility we have that $h,h' \in H$ Let $h, h' \in H$ . Let $(\gamma_1,
  \gamma_2) := (\gamma_1', \gamma_2') := (\alpha, \beta)$. This assignment is
  sufficient to prove the direction.

  $`\Leftarrow'$: Suppose for all $h,h' \in H$ we have $(\gamma_1, \gamma_2),
  (\gamma_1, \gamma_2') \in \sym_A \times \sym_B$ satisfying the above
  requirements. Notice that for a given $i \in A$ and any $j, j' \in B$, let $h
  = \omega(i,j)$ and $h' = \omega(i,j')$, then we have that $\gamma_1 (i) =
  \gamma_1'(i)$. It follows that we can define a $\alpha \in \sym_A$ by
  $\alpha(i) = \gamma_1 (i)$. Similarly we can define $\beta \in \sym_B$ by
  $\beta (j) = \gamma_2 (j)$.

\end{proof}

\begin{definition}
  Let $g$ be a gate with matrix labeling $\omega$. Let $(\sigma, h, h') \in
  \sym([n]) \times H^2$. Say that $(\sigma, h, h')$ is useful if $(h,h')$ is
  incompatible with $(\sigma, \omega)$.

  Say that two distinct pairs $(\sigma_1, h_1, h_1'), (\sigma_2, h_2, h_2') \in
  \sym([n]) \times H^2$ are mutually independent if
  \begin{itemize}
  \item $\sigma_2 h_1 = h_1$,
  \item $\sigma_2 \sigma_1 h_1 = \sigma_1 h_1$,
  \item $\sigma_2 h_1' = h_1'$,
  \item $\sigma_2 \sigma_1 h_1' = \sigma h_1'$,
  \end{itemize}
  We say that a set $S \subseteq \sym([n]) \times H^2$ is useful if each pair in
  it is useful. We say that S is independent if each pair of distinct pairs in
  $S$ are mutually independent.
\end{definition}

We denote the usual equivalence relation on $\sym_n$ from the (right) cosets of
$\matstab(\omega)$ by $\sim_\omega$.

\begin{lem}
  Let $\sigma_1, \sigma_1 \in \sym_n$ and suppose $\sigma_1 \sim_\omega
  \sigma_2$ then for any $(h,h') \in H^2$ we have that $(h,h')$ is compatible
  with $(\sigma_1, \omega)$ iff $(h,h')$ is compatible with $(\sigma_2,
  \omega)$.
\end{lem}

\begin{proof}
  Suppose we have that $\sigma_1 \sim_\omega \sigma_2$. And additionally
  supposed $(h, h')$ compatible with $(\sigma_1, \omega)$.

  It follows there exists $(\alpha, \beta) \in \sym_A \times \sym_B$ such that
  $\sigma_1 \omega (i,j) = \sigma_2\omega (\alpha i, \beta j)$, and we have
  $(\gamma_1, \gamma_2), (\gamma_1', \gamma_2') \in \sym_A \times \sym_B$
  satisfying the requirements of compatibility. We note that $(\alpha, \beta)
  \cdot \omega^{-1}(\sigma_1 h) = \omega^{-1}(\sigma_2 h)$. Thus by composition
  we have that $(\alpha \gamma_1, \beta \gamma_2) \cdot \omega^{-1}(h) =
  \omega^{-1}(\sigma_2 h)$ and $(\alpha \gamma_1', \beta \gamma_2') \cdot
  \omega^{-1}(h') = \omega^{-1}(\sigma_2 h')$, and clearly the remaining
  requirement for compatibility follows from the bijectivity of $\alpha$ and
  $\beta$. The result follows. The other direction of the implication follows
  from symmetry.
\end{proof}

\begin{claim}
  \label{claim:useful-independant-set}
  Let $g$ be a rank gate with labeling $\omega$ and child set $H$. Let $S$ be a
  useful and independent. We then have that $\vert \sym_n: \matstab(\omega)
  \vert \leq 2^{\vert S \vert}$.
\end{claim}

\begin{proof}
  Let $R \subseteq S$ and define $\sigma_R = \Pi_{(\sigma, h, h') \in R} \sigma$
  (with some arbitrary order assumed on $S$). Let $R$ and $Q$ be distinct
  subsets of $S$ and WLOG let $\vert R \vert \geq \vert Q \vert$. We want to
  show that we don't have $\sigma_R \omega \sim_\omega \sigma_Q \omega$. Pick
  any $(\sigma, h, h') \in R/Q \neq \emptyset$. Given that $\sigma_R h = \sigma
  h$ and $\sigma_R h' = \sigma h'$ it is easy to see that the usefulness of
  $(\sigma, h,h')$ implies the incompatibility of $(h,h')$ with with $(\omega,
  \sigma_R)$. Moreover, the fact that $\sigma_Q h = h$ and $\sigma_Q h' = h'$
  makes it easy to see $(h,h')$ is compatible with $(\omega, \sigma_Q)$. From
  the above lemma we may conclude that that we do not have $\sigma_R \sim_\omega
  \sigma_Q$, and the result follows.
\end{proof}

The following two lemmas proved by Anderson and Dawar \cite{} are both of use in
proving the following theorem.

\begin{lem}
  \label{lem:big-or-small}
  For any $\epsilon$ and $n$ such that $0 < \epsilon < 1$ and $\log n \geq
  \frac{4}{\epsilon}$, if $\mathcal{P}$ is a partition of $[n]$ with $k$ parts,
  $s = [\sym_n : \setstab (\mathcal{P})]$ and $n \leq s leq 2^{n^{1-\epsilon}}$,
  then $\min \{k, n-k\} \leq \frac{8}{\epsilon} \frac{\log s}{\log n}$.
\end{lem}

\begin{lem}
  \label{lem:small-means-support}
  For any $\epsilon$ and $n$ such that $0 < \epsilon < 1$ and $\log n \geq
  \frac{8}{\epsilon^2}$, if $\mathcal{P}$ is a partition of $[n]$ with $\vert
  \mathcal{P} \vert \leq \frac{n}{2}$, $s:= [\sym_n : \setstab (\mathcal{P})]$
  and $n \leq s \leq 2^{n^{1-\epsilon}}$, then $\mathcal{P}$ contains a part $P$
  with at least $n - \frac{33}{\epsilon} \cdot \frac{\log s} {\log n}$.
\end{lem}

If $g$ is a symmetric gate (i.e. the usual gates in a circuit) we note that
$\orb(g) = [\sym_n : \stab (g)]$ by the orbit-stabilizer theorem.

If $g$ is a matrix-symmetric gate then $\orb(g) = [\sym_n:\matstab (\omega)]$,
where $\omega$ is the matrix labelling associated with $g$.

\begin{thm}
  For any $\epsilon$ and $n$ such that $\frac{2}{3} \leq \epsilon \leq 1$ and $n
  \geq \frac{128}{\epsilon^2}$, if $C$ is a symmetric, rigid circuit on
  structures of size $n$ and $s := \max_{g \in C} \vert \orb (g)\vert \leq
  2^{n^{1-\epsilon}}$, then, $SP(C) \leq \frac{33}{\epsilon}\frac{log s}{log
    n}$.
\end{thm}

\begin{proof}
  It is easy to see that if $g$ is a gate in $C$ then $\stab (g) \subseteq
  \setstab(\SP(g))$, and so $s \geq \orb(g) = [\sym_n : \stab(g)] \geq [\sym_n :
  setstab(\SP(g))]$. Thus if $\vert \SP(g) \vert \leq frac{n}{2}$, then from
  Lemma \ref{lem:small-means-support}, we have $\| SP (g) \| \leq
  \frac{33}{\epsilon} \cdot \frac{\log s} {\log n}$. The result thus follows
  from showing that for each $g$ in $C$ we have that $\vert \SP (g) \vert \leq
  \frac{n}{2}$.

  % Include some detail here for constant and relational gates
  The cases where $g$ is a constant or relational gate are easy to handle.

  We now consider the case for internal gates. Let $g$ be the topologically
  first internal gate with $\vert \SP(g) \vert > \frac{n}{2}$. If $g$ is not a
  matrix-symmetric gate then the result follows from the argument presented by
  Anderson and Dawar \cite{}. Suppose $g$ is a matrix-symmetric gate, and
  suppose $g$ has a labelling $\omega$. We now argue that this leads to a
  contradiction.

  Let $k' := \lceil \frac{8 \log s}{\epsilon \log n} \rceil$. From the
  assumptions on $s, n$ and $\epsilon$ we have that $k' \leq
  \frac{1}{4}n^{1-\epsilon} < \frac{n}{2}$. Lemma \ref{lem:big-or-small} implies
  that $n - \vert \SP(g) \vert \leq k'$

  From Claim \ref{claim:useful-independant-set} it remains to show that we can
  construct a sufficiently large useful and independent set of gate-automorphism
  pairs $S$. Divide $[n]$ into $\lfloor \frac{n}{k' + 2} \rfloor$ disjoint sets
  $S_i$ of size $k' + 2$ and ignore the elements left over. It follows that for
  each $i$ there is a permutation $\sigma_i$ which fixes $[n] / S_i$ pointwise
  but moves $g$. Suppose there was no such $\sigma_i$ it follows that every
  every permutation that fixes $[n]/S_i$ pointwise fixes $g$. Thus the partition
  of all the singletons in $[n]/S_i$ and $S_i$ is a supporting partition. As
  $\SP(g)$ is the coarsest partition it follows that $\vert \SP(g) \vert \leq n
  - (k'+2) + 1 = n - k' - 1$, which contradicts the inequality $n - \vert \SP(g)
  \vert \leq k'$.

  Since $g$ is moved by each $\sigma_i$ and $C$ is rigid it follows that we
  don't have $\sigma_i \notin \matstab(\omega)$. Thus there exists $(h_i, h_i')$
  that is inconsistent with $(\sigma_i, \omega)$, and so the triple $(\sigma_i,
  h_i, h_i')$ is useful.

  Let $\SP (h)^*$ be the union of all parts of $\SP(h)$ except for the largest
  part. Let $Q_i = \SP(h_i)^* \cup \SP(\sigma_i h_i)^* \cup \SP (h_i')^* \cup
  \SP (\sigma_i h_i')^*$. Then note that if $\sigma_j$ fixes $Q_i$ then by
  construction, we have that $\sigma_j \in \stab_n(\SP(h_i)) \cap
  \stab_n(\SP(\sigma_i h_i)) \cap \stab_n(\SP(h_i')) \cap \stab_n(\SP(\sigma_i
  h_i'))$

  Define a graph $K$ with vertices given by the sets $S_i$ and an edge from
  $S_i$ to $S_j$ (with $i \neq j$) if $Q_i \cap S_j \neq \emptyset$. It follows
  then that if there is no edge between $S_i$ and $S_j$ then $(\sigma_i, h_i,
  h_i')$ and $(\sigma_j, h_j, h_j')$ are mutually independent. It remains to
  argue that $K$ has a large independent set. This is possible as the out-degree
  of $S_i$ in $K$ is bounded by
  \begin{align*}
    \vert Q_i \vert \leq \|SP(h_i) \| + \|SP(\sigma_i h_i) \| + \|SP(h_i') \| + \|SP(\sigma_i h_i') \leq 4 \cdot \frac{33\log s}{\epsilon \log n}
  \end{align*}.

  This follows as the sets $S_i$ are disjoint and we may apply Lemma
  \ref{lem:small-means-support} to each of the child gates. It follows that the
  average total degree (in + out degree) of $K$ is at most $2 \cdot \vert Q_i
  \vert \leq 34 \cdot k'$. Now greedily select a maximal independent set in $K$
  by repeatedly selecting $S_i$ with the lowest total degree and eliminating it
  and its neighbours. This action does not affect the bound on the average total
  degree of $K$ and hence determines an independent set $I$ in $K$ of size at
  least
  \begin{align*}
    \frac{\lfloor \frac{n}{k' + 2} \rfloor}{34k' + 1} \geq \frac{n - (k'+2)}{34k'+1k'+2} \geq \frac{n\frac{7}{16}}{34k'^2 + 69k' +2} \geq \frac{n}{(16k')^2}.
  \end{align*}

  Take $S = \{(\sigma, h, h') : S_i \in I \}$. Then from the above argument we
  have that $S$ is useful and independent.

  Moreover, from Claim \ref{claim:useful-independant-set}, we have that $s \geq
  \vert \orb(g) \vert \geq 2^{\vert S \vert} \geq 2^{\frac{n}{(16k')^2}}$ then
  $n^{1-\epsilon} \geq \log s \geq n \cdot (\frac{128}{\epsilon}\frac{\log
    s}{\log n})^{-2} > n \cdot (n^{1-\epsilon})^{-2} = n^{2\epsilon -1} \geq
  n^{1-\epsilon}$. This is a contradiction.
\end{proof}

\section{Rigid Circuits}

\begin{definition}
  We say that a circuit $C$ has bijective labels if for each gate $g$ in $C$,
  $L(g)$ is a bijection.
\end{definition}

\begin{lem}
  \label{lem:bij_labels}
  There is an algorithm that runs in polynomial time that takes in a circuit $C$
  and outputs a circuit with unique gates $C'$. Moreover, if $C$ was symmetric
  then $C'$ is symmetric. If $C$ is rigid then $C'$ is rigid.
\end{lem}

\begin{proof}
  Let $S = 2*\vert C \vert$. Recurse through the gates of $C$ topologically and
  let $h$ be the next gate topologically. If for all $g \in W(h, \cdot)$ we have
  that $L\vert (g)^{-1}(h) \vert = 1$ continue on to the next gate. If not add
  in a tower of $S$ $\and$ gates such that $h \rightarrow \and^h_1 \rightarrow
  \ldots \rightarrow \and^h_S$ (i.e. we have a tower of $\and$ gates with $h$ as
  input to $\and^h_1$ and the output of each $\and^h_i$ connected to the input
  of each $\and^h_{i+1}$ for each $1 \leq i < S$). Now for each $g \in W(h,
  \cdot)$, if $L(g)^{-1}(h) = \{ s_0, \ldots, s_{r}\}$, for each $1 \leq i \leq
  r$ add in the wires $W(\and^h_i, g)$ and set $L(g)(s_{i}) = \and^h_i$. Now
  continue on to the next gate topologically and run the above algorithm.

  % First, notice that for a given gate $h$ and $k \in \mathbb{N}$, we can
  % define
  % a sub-circuit $h^k = 1 \and \cdots, 1 \and h$ (or rather a $k$-height tower
  % of
  % $k$ binary $\and$ gates with $h$ at the top and all other inputs set to
  % $1$).
  % We call $h$ the top of the sub-circuit $h^k$ and the bottom $and$ gate (the
  % output of the sub-circuit) the bottom gate. We may think of this as a
  % $k$-height copy of the gate $h$, in the sense that it has the same output as
  % $h$ and similar orbits. We now use different copies to distinguish gates
  % that
  % otherwise have the same labelling.

  % Let $h$ be the next gate topologically. Then let $r$ be maximal such that
  % $W(h,g)$ and $\omega_g^{-1}(h) = \{ s^g_1, s^g_2, \ldots, s^g_r \}$. Then
  % let
  % $k$ be the height of the highest tower with $h$ at the top and with it's
  % bottom connected to a gate in $W(h, \cdot)$. Then create a single
  % $k+r-1$-height tower by adding in the appropriate number of binary $\and$
  % gates below $h$.

  % For each $g \in W(h, \cdot)$ and for each $\omega_g^{-1}(h) = \{s^g_1,
  % s^g_2,\ldots , s^g_{r_g}\}$, set $L'(g)(s^g_1) = h$ each child starting with
  % the $and$-gate child to $h$ add in a wire from from the $i$th gate in the
  % tower to $g$ and set $L'(g)(s^g_{i+1}) = \and_i$, where $\and_i$ is the
  % $i$th
  % $\and$ gate in the tower starting with the $\and$ gate in the tower child to
  % $h$. There are enough gates in the tower as $r_g \leq r$.

  We call this updated circuit $C'$.

  Firstly, note that after running the above algorithm for each $g$ the
  labelling of $g$ will be a bijection. Moreover, it's easy to see that the
  output of each gate remains unchanged, and as such the output of the circuit
  is unchanged.

  Secondly, notice that the size of $C'$ is at most $2*\vert C \vert^2$, and
  note that the above algorithm runs in polynomial time.

  Now suppose that $C$ is symmetric. Let $\sigma$ be a permutation on the input
  universe and $\pi_C$ the induced automorphism on $C$. We now define $\pi$, the
  induced automorphism on $C'$. For each gate $g$ in $C'$, if $g$ is in $C$ then
  set $\pi(g) := pi_C(g)$. If $g$ is not in $C$ then $g$ must be some
  $\and^h_i$, for some $h$ in $C$. Then set $\pi(\and^h_i) :=
  \and^{\pi_C(h)}_i$. It is easy to see that $\pi$ is an automorphism, and $\pi$
  extends $\sigma$.

  Suppose that $C$ is rigid. It is easy to see that $C'$ will be rigid as well.

\end{proof}

% If $h$ is a binary $\and$ gate, then we say that $h$ has height $k+1$ if it
% has a $k$-height tower of binary $and$ gates connected above it, with a
% non-$\and$ gate $p$ connected to the final binary $\and$ gate, and all other
% inputs connected to $1$. We then say that $h$ is the base of a tower of height
% $k$ and type $p$.

% Then for each $h \in W_g$, if $h$ is not a binary $\and$ gate, the let $k$ be
% the largest integer such that there a gate $h' \in W(h, \cdot)$ which is the
% base of a tower of height $k$ and type $h$. it's output directly connected to
% one of the gates in $W(h^k, \cdot)$ (with k = 0 corresponding to just $h$).
% Then each $i \in \{1, \ldots , \vert \omega^{-1}(h)\vert\}$ add into the
% circuit $h^{k+i}$ and wire it to $g$. Notice that this step requires adding
% $\vert \omega^{-1}(h) \vert -1$ gates to the circuit. For some assumed
% ordering $\dom(\omega)$, let $\{s_0, \cdot, s_m\} = \omega^{-1}(h)$. Define
% $L'(g)(s_i) = h^{k+i}$. Now continue on to the next gate topologically where
% $L(g)$ is not a bijection, and execute the same process

\begin{lem}
  There is an algorithm that runs in polynomial time that takes in a circuit $C$
  and outputs a circuit $C'$ such that $C'$ is rigid and has bijective labels.
  Moreover, if $C$ was symmetric it follows that $C'$ will be symmetric.
\end{lem}

\begin{proof}
  First run the algorithm from Lemma \ref{lem:bij_labels} on $C$, and call the
  output circuit $C$.

  Recurse through the gates of $C$ topologically. For each internal gate $g$,
  for all $g'$ in $C$ such that $g \neq g'$, $W(\cdot, g) = W(\cdot, g')$, $W(g,
  \cdot) = W(g', \cdot)$, $\Sigma(g) = \Sigma(g')$, $L(g) \sim L(g')$ and
  $\Omega^{-1}(g) = \Omega^{-1}(g')$, delete $g'$ and for all $s \in L^{-1}(g')$
  set $L(s) := g$.

  Now re-run the algorithm from Lemma \ref{lem:bij_labels} on $C$ and output the
  result.
\end{proof}

  \begin{lem}
    Let $C = \langle G, W, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\SB, \MB,
    \tau)-circuit$ on structures of size $n$. There is a deterministic algorithm
    which runs in Poly($\vert C \vert$) and outputs a rigid $(\SB, \MB, \tau)
    circuit$ $C'$ such that $G' = G$ and for any $g \in G$, and any input
    $\tau$-structure $\mathcal{A}$ and any bijection $\gamma$ from $A$ to $[n]$,
    $C[\gamma \mathcal{A}](g) = C'[\gamma \mathcal{A}](g)$ and if $C$ is
    symmetric then so is $C'$.
  \end{lem}

  \begin{proof}

  \end{proof}

  \section{Computing Supports}
  \begin{lem}
    Let $C$ be a rigid $(\SB, MB, \tau)$-circuit on structures of size $n$ and
    $\sigma \in \sym_n$. There is a deterministic algorithm which runs in time
    Poly($\vert C \vert$) and outputs for each gate $g$ its image under the
    automorphism $\pi$ induced by $\sigma$, if it exists.
  \end{lem}
  \begin{proof}
    The proof proceeds by recursively going through the circuit and building the
    mapping $\pi$ induced by $\sigma$.

    Suppose $g$ is a constant gate, then $\pi g := g$. Suppose $g$ is a
    relational gate, then there is at most one gate $g'$ such that $\Sigma (g) =
    \Sigma (g')$ and $\sigma\Lambda (g') = \Lambda (g)$. If such a $g'$ exists
    assign $\pi g := g'$, else terminate with failure.

    If $g$ is an symmetric internal gate then (from rigidity) there is at most
    one gate $g'$ such that $\Sigma (g) = \Sigma(g')$ and $W_{g'} = \pi W_g$.
    Assign $\pi g := g'$ if such a gate exists, or else terminate with failure.

    If $g$ is a matrix-symmetric internal gate then consider the set of gates
    $g'$ such that $g'$ has children $\pi W_g$ and $\Sigma(g) = \Sigma(g')$, and
    let $A \times B = \dom (Sigma(g))$. If no such gate $g'$ exists, terminate
    with failure. Define $\sigma_{\pi, g'}:A \times B \rightarrow A \times B$ by
    $\sigma_{\pi, g'} = \omega^{-1}_{g'} \pi \omega_{g}$. Then clearly
    $\omega_{g'} \sigma_{\pi, g'} = \pi \omega_{g}$, and it's easy to show that
    $\pi \omega_g \sim \omega_{g'}$ iff $\sigma_{\pi,g'} \in \sym_A \times
    \sym_B$. But this just involves checking that $\sigma$ acts as a bijection
    on $A$ and $B$ separately and, given that $\vert A \vert$ and $\vert B
    \vert$ are both bounded by $\vert C \vert$, the algorithm which just
    iterates through $A$ and $B$ is sufficient. If for every $g'$ it is found
    that $\pi_{\sigma,g'}$ is not in $\sym_A \times \sym_B$ then terminate with
    failure. If there is a $g'$ for which $\pi_{\sigma, g'} \in \sym_A \times
    \sym_B$ then it is unique by rigidity and so set $\pi g := g'$.

    If $g$ is an output gate, then check that for all tuples in $[n]^{q}$ we
    have that $\pi \Omega (x) = \Omega (\sigma (x))$, and terminate with failure
    if the condition is not met.

    If the algorithm has not terminated with failure, output the automorphism.

    The algorithm clearly runs in Poly($\vert C \vert$)
  \end{proof}

>>>>

\documentclass[12pt]{article}

% This first part of the file is called the PREAMBLE. It includes customizations
% and command definitions. The preamble is everything between \documentclass
% and \begin{document}.

\usepackage[margin=1in]{geometry} % set the margins to 1in on all sides
\usepackage{graphicx} % to include figures
\usepackage{amsmath} % great math stuff
\usepackage{amsfonts} % for blackboard bold, etc
\usepackage{amsthm} % better theorem environments
\usepackage{xspace}
\usepackage{subfiles}%allow for subfiles
\usepackage{blindtext}
\usepackage{mymacros}

\newtheorem{thm}{Theorem}
\newtheorem{claim}[thm]{Claim}
\newtheorem{remark}[thm]{Remark}
\newtheorem{definition}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{conj}[thm]{Conjecture}

@@

\begin{document}

\title{Symmetric Circuits for Rank Gates}

\author{Anuj Dawar and Gregory Wilsenach \\ 
Computer Laboratory \\
University of Cambridge}

@@

\tableofcontents
\newpage

\section{Introduction}
\subfile{sections/introduction}
\newpage

\section{Background}
\subfile{sections/background}
\newpage

\section{Symmetric Circuits}
\subfile{sections/symmetric-circuits}
\newpage

\section{Symmetry and Support}
\subfile{sections/symmetry-and-support}
\newpage

\section{Translating Circuits into Formulas}
\subfile{sections/translating-circuits-into-formulas}
\newpage

\section{Concluding Remarks}
\subfile{sections/conclusion}
\newpage

\bibliographystyle{plain} \bibliography{references}
\end{document}
